Multi-scale trend analysis of water quality using error propagation of generalized additive models

### 2021-02-04 manu_draft.Rmd
working through methods for coauth comments

Accurate and flexible trend assessment methods are valuable tools for describing historical changes in environmental monitoring datasets.  A key requirement is complete propagation of uncertainty through the analysis.  However, this is difficult when there are mismatches between time scales of monitoring data and analysis methods. Here, we propose a novel application of Generalized Additive Models (GAMs) to model seasonal and multi-decadal changes in a long-term monitoring dataset of chlorophyll-a concentrations in South San Francisco Estuary.  GAMs are a general method for modelling a response variable as the additive sum of one or more smoother splines fit to predictor variables.  These models have shown promise in other applications to water quality monitoring data to separate long-term (i.e., annual or decadal) trends from seasonal variation.  Our proposed methods estimate seasonal averages in a response variable with GAMs and then use the uncertainty measures with mixed-meta regression analyses to quantify inter-annual trends that account for full propagation of error across methods.  We first demonstrate that nearly identical descriptions of temporal changes can be obtained using different smoothing splines for annual or seasonal components of the time series.  We then extract seasonal averages for an a priori period of time within each year from the GAM results, including an accurate assessment of standard error within each year where the seasonal average is estimated.  Finally, we demonstrate how trends can be quantified with mixed-meta regression analyses that evaluate changes in the seasonal averages between years with full propagation of uncertainty in the estimates.  Overall, this approach leverages the ability of GAMs to identify an optimal amount of smoothing from the raw data and uses those results to accurately assess trends that minimize the likelihood of false positive or negative results, which is particularly important for time series with missing observations or varying sample effort over time.

@@

We evaluated near-surface chlorophyll-a data measured biweekly to monthly from 1990 to 2019 along the longitudinal axis of the SFE extending from Central Bay (stations 18-23), South Bay (stations 24-32), and Lower South Bay (stations 34-36) (Table \@ref(tab:sumtab), Figure \@ref(fig:sitemap)). Monitoring data were obtained from the SFE Research Program of the US Geological Survey [@Cloern16;@Schraga20].  Chlorophyll on GFF filters was analyzed fluorometrically after extraction in 90% acetone. Sampling frequency varied somewhat over time and by station.  Every observation was included directly in the statistical models without spatial or temporal binning or averaging.

@@

where `y` is the time-series of interest, such as chlorophyll, `cont_year` is "continuous year", a continuous numerical date (e.g., July 1st 2019 would be 2019.5), `y ~ s(...)` indicates that `y` will be explained by a smoothing spline (in this case of `cont_year`), and `num_knots_Y` is the number of knot or "connections" along the spline that influence curvature.  Log$_{10}$-transformed chlorophyll was used for all analyses to meet assumptions of normally-distributed residuals.

@@

Model SYD has the appealing feature that, if some parts of some years have limited data, model SYD will consider data from the same period in other years by imputing an average seasonal pattern with the `doy` spline. However, an interpretation of these imputations is challenging.  For example, the spring chlorophyll peak is a notable feature every year in south SFE.  If the peak occurs at the same time every year but the magnitude varies, then the average within-year pattern can be interpreted as the average magnitude. However, if the magnitude is the same but the timing varies across years, then the magnitude of the average peak cannot be similarly interpreted and instead underestimates the magnitude that usually occurs.  Moreover, the width or duration of the peak will be longer than typically occurs in a given year.

@@

The distinct aspect of model SYDI is the anticipation that within-year fluctuations will vary smoothly from year to year.  For example, if one year has an especially early and large magnitude spring bloom, the years before and after would be expected also to have earlier and larger than average spring blooms. For the raw SFE data, and chlorophyll-a dynamics in many estuaries, this is not expected to be the case. An early and large spring bloom could easily be followed the next year by a late and relatively small spring bloom.  If the interaction spline has many knots, then the degree of smoothness across years can be estimated to be very low, allowing fluctuations within years to vary strongly across years.  However, if that is the case, the conceptual motivation for model SYDI may not be supported by the data. Given that observed bloom dynamics for most systems do not vary smoothly across years, the potential benefits of using model SYDI could be outweighed by using simpler models.  Moreover, the spline and penalty structure differ from the other models, making it difficult to evaluate their relative performance (Table \@ref(tab:modstrtab)).

@@

We visually compare chlorophyll estimates from models SY, SYD, and SYDI to emphasize that similar fits can be achieved by all of the presented models (Figure \@ref(fig:modsumfig), S is identical to SY and is not shown).  Models SY, SYD, and SYDI were fit to chlorophyll data from station 34 using large k values for the arguments `num_knots_y`, `num_knots_D`, `num_knots_Y_ti`, and `num_knots_D_ti` for each model.  Predictions by day of year from each model are visually similar (Figure \@ref(fig:modsumfig)a) and closely follow the 1:1 line (Figure \@ref(fig:modsumfig)b).  However, when contrasting the models using only the continuous year smoother (`s(cont_year)`, as an example), the model fits differ substantially because of structural differences and differences in the penalties applied to the basis functions.  These results are also reflected in differences in the effective degrees of freedom among the additive components of each model (Table \@ref(tab:modstrtab)). Accordingly, even though the models differ by which structural component describes variation in the chlorophyll time series, they provide similar predictions.

For all results, model S was used with enough knots in `num_knots_y` to evaluate chlorophyll trends across the monitoring stations in South SFE.  This model was chosen because of the relatively quicker processing time to fit the model, while providing nearly identical explanatory power as compared to the other models (Table \@ref(tab:modstrtab)).

@@

In the second-stage analysis, we estimated a seasonal average, such as the mean spring chlorophyll concentrations and the uncertainty in each year.  Define $\mu_t$ as the average for the period of interest (e.g., seasonal average) in year $t$, $\hmt$ as an estimate of $\mu_t$, and $\hshmt$ as the estimated standard error of $\hmt$. The season includes $n$ days.  For simplicity, the following text omits subscript $t$.

Point estimates of response values for the fitted GAM take the form $\hby = \bX \hbb$, where $\hbb$ is the vector of parameter estimates and $\mathbf{X}$ is a model matrix of explanatory variables, including spline basis function values.  Vector $\hbb$ includes both fixed effect parameters and spline parameters, and $\bX$ contains columns corresponding to each.  For example, using model SY, if a point estimate for chlorophyll is needed for a single day, given as `dec_year` = $r$, then $\bX$ would have a row with $1$ in the first column (for the intercept parameter), $r$ (for the linear time trend) in the second column, and an evaluation of each spline basis function at $r$ (and possibly its corresponding `doy` value) in the remaining columns. The number of spline basis functions is related to the number of knots. Note that $r$ can be any time, not necessarily the time of an observation.

@@

In stage three of the analysis, we used a meta-analysis method to evaluate across-year trends of within-year features such as seasonal water quality, characterized by the within-year means ($\hmt$) and their standard errors ($\hshmt$) that we estimated in stage two of the analysis.  In the examples, we considered two periods of interest for chlorophyll, January-July and August-December, which are relevant to phytoplankton bloom phenology in south SFE.

This analysis provided a direct answer to the question: Is there a significant linear trend across a group of years in a seasonal average, where the time-scale of the trend is chosen by the investigator?  For example, is there a trend in the spring chlorophyll average from 1990 to 2000? This question can also be posed in a moving-window manner across a time-series (e.g., spring average trend from 1990-2000, 1991-2001, etc.). For all analyses, the response data of interest are $\hmt$, $t = 1, \ldots, N$, with their associated standard errors, $\hshmt$.  $N$ is the number of years of the study.

### Linear trends across years at a chosen time-scale

@@

This method was applied to a chosen sequence or "window" of years for estimating the linear trend.  This approach evaluated whether there was a significant linear trend and provided an estimated rate of change (slope) in chlorophyll over a chosen series of years. Using the meta-analysis considered statistical properties derived from all the data in the years of interest (i.e., the seasonal averages and their uncertainty).  Years with more data have smaller standard errors for $\hmt$ compared to years with less data. The meta-analysis explicitly includes these differences in uncertainty magnitudes.

@@

The above methods were applied to each station by evaluating changes in seasonal averages from January to June and July to December for approximate ten year blocks from 1991 to 2000, 2000 to 2010, and 2010 to 2019. As an extension of these methods, a chosen sequence of years was applied across the time series to assess periods of time within which seasonal trends could be significant or not (e.g., spring average trend from 1990-2000, 1991-2001, etc.).  Although the initial year sequence and seasonal period to evaluate is chosen by the analyst, applying the method in a moving window approach reduces some of the ambiguity around points in the time series when trends may be changing with reasonable certainty.  The moving window approach applies the meta-analysis model from left to right in the time series across the seasonal averages to obtain the slope and significance of the estimated trend.  We chose a centered five-year window where the model estimates are based on results at equal half-window widths to the left and right of a given year. Although a left- or right-centered window could also be applied, we limit the analysis to a centered window to demonstrate the concept.

@@

where the back-transformed, expected value of the response variable $y$ (chlorophyll) is a function of the mean value $\mu$ in log$_{10}$-space and a dispersion estimate $\sigma^2$ from the model.  The dispersion is obtained directly from the fitted GAM object as an estimate of the variance of the residuals divided by the model degrees of freedom.

Slope estimates that were used to determine significance of a seasonal trend across years from the mixed-meta analysis were not back-transformed.  While it is possible to back-transform model predictions, estimates of the mean values, and endpoints of confidence intervals, it is not possible for slope estimates in log$_{10}$-space that can change in arithmetic space across the values of the independent variable.  As such, all results reporting slope estimates as rates of change in chlorophyll per year were reported at the scale of the model in log$_{10}$-space.

@@

Model predictions for chlorophyll trends across all stations had an average R-squared value equal to `r aver2`%  (Table \@ref(tab:modprftab)) and ranging from `r minr2`% (station `r minst`) to `r maxr2`% (station `r maxst`). GAM predictions from north to south on the longitudinal axis showed more pronounced annual and seasonal changes in chlorophyll towards the more southern stations (Figure \@ref(fig:prddat)). All the models suggested 1) increasing chlorophyll from 1990 until 2005 to 2010, followed by decreasing chlorophyll until the end of the record in 2019, 2) a spring chlorophyll peak, particularly at South Bay stations, and 3) a fall chlorophyll peak that was smaller than the spring peak.  The magnitude of the fall peak did not vary noticeably by location (Figure \@ref(fig:prddat)).

@@

To demonstrate the differences in results that can be obtained with the trend tests, estimates of linear trends across years for different seasons are shown for station 34 (Figure \@ref(fig:trnddat)). All plots show trends in seasonal averages within ten-year windows, whereas rows a-c show estimates from January to June and rows d-f show estimates from July to December.  The seasonal trend analyses showed that January to June chlorophyll increased (log$_{10}$ Chl-a slope 0.03 $\mu$g L$^{-1} yr$^{-1}$, 0.01-0.06 95\% confidence interval) form 1991 to 2000, whereas a trend for the same period in July to December was not observed.  Chlorophyll also increased from 2000 to 2010, but only for July to December (log$_{10}$ slope 0.03, 0.01-0.05 95\% confidence interval).  Finally, chlorophyll decreased from 2010 to 2019 but only for July to December (log$_{10} Chl-a slope -0.02, -0.04-0 95\% confidence interval).  Because the trends were confined to certain times of the year, the seasonal estimates provide additional information beyond coarser estimates that cover the entire year.

Temporal changes varied among regions of the Bay and were in fact estimated to be moving in opposite directions.  Figure \@ref(fig:trndmap) shows results from similar analyses as those in figure \@ref(fig:trnddat), but applied to all stations.  
Mixed-meta regression analyses applied to seasonal averages showed that increases (based on $p < 0.05$) for the January to June period were observed at stations 32, 34, and 36 from 1991 to 2000 and station 18 from 2000 to 2010, whereas chlorophyll decreased for at stations 30 and 32 from 2010 to 2019. For the July to December period, increases were observed at stations 24, 27, 30, and 32 from 1991 to 2000 and stations 18, 21, 22, and 34 from 2000 to 2010, whereas decreases were observed at stations 30, 32, and 34 from 2010 to 2019.

@@

Results from a five-year moving window comparison of seasonal trends provided additional context on when significant changes were occurring at each station (Figure \@ref(fig:winchg)).  The largest trends occurred at southern stations where increases occurred in both halves of the year (i.e., January to June, July to December).  Decreases were observed later in the record, until recent years when trends were  Significant increases were observed for both January to June and July to December earlier in the period of record, whereas significant decreases were observed later in the record until chlorophyll was relatively stable across all stations and seasons. A change from increasing to decreasing chlorophyll occurred at most stations around 2007, particularly for January to June.

@@

Applying the same comparison to all stations showed that different trend analysis methods provided conflicting information on the magnitude and significance of the seasonal chlorophyll changes in each decade (Figure \@ref(fig:trndcmp)).  The slope estimates from the OLS models applied to the observed data were understandably more variable than the slope estimates from the GAM averages and mixed-meta methods, with much larger slopes observed especially at the more southern stations in the January to March period.  Slope estimates from the OLS models applied to the averages from the GAMs as compared to the mixed-meta results were more similar, excluding some of the slope estimates for the southern stations.  Differences in significance of trends between the OLS models applied to the GAM averages and the mixed-meta analyses were also observed, reflecting the ability of the former to more accurately assess significance of trends by accounting for uncertainty in the average estimates.

@@

The trend assessment approach does not address causality in the observed trends.  In this regard, the approach is like other trend assessment methods where the focus is on understanding direction, magnitude, and confidence in changes in water quality variables over time. However, a logical follow-up question is "what" is driving the trend after the trend has been adequately described.  This information has obvious implications for management decisions on factors that influence water quality changes, e.g., wastewater treatment upgrades, large-scale climatic factors, or flow regulation practices.  An advantage of GAMs is their flexibility in including alternative predictors, such that the significance of a predictor or comparison of nested models with and without different predictors can provide evidence of which predictors are driving the observed trends [@Wood02;@Zuur09]. In such cases, considerations of model structure can have direct implications on conclusions regarding potential causality.  As noted above, model structure in describing the long-term trend component was irrelevant for describing long-term trends, although a distinction here is made in the objective of the analysis.  Our goal herein was to describe chlorophyll changes relative to time where the predictors were variations on a general theme (e.g., season vs. year), whereas using GAMs with different predictors assess potential causality is a different application with alternative goals.  Therefore, using our approach to assess causality will require considerations of model structure given how GAMs could be used to assess different questions.

Finally, the evaluation of trends for alternative water quality variables in addition to chlorophyll is a simple and logical extension of the methods proposed in this study.  The long-term monitoring program maintained by USGS includes multiple parameters in addition to chlorophyll that can provide additional context into broader water quality trends in South Bay [@Cloern16;@Schraga20]. These parameters include salinity, temperature, light attenuation, dissolved oxygen, suspended particulate matter, and dissolved inorganic nutrients, which collectively can be used to provide a broader understanding of potential eutrophication patterns or ecosystem shifts at seasonal and multi-decadal scales.  Chlorophyll measurements can also be used to estimate gross primary production to assess process rates that may be more indicative of system function [@Jassby02;@Cloern07].  The open-source R package developed for this manuscript can be used for these analyses to provide additional insight into potential drivers of water quality change in SFE and other estuarine systems.

@@

ylab <- expression(paste("Chlorophyll-a (", italic(mu), "g ", L^-1, ")"))

@@

p <- p1 + p2 + plot_layout(ncol = 1)
png('figs/obsdat.png', height = 7, width = 8, family = 'serif', units = 'in', res = 400)
print(p)
dev.off()
```
```{r obsdat, fig.cap = "Observed chlorophyll concentrations for all stations in central and south San Francisco Estuary (18-36, Figure \\@ref(fig:sitemap)), with (a) annual summer/fall concentrations (Aug - Dec) and (b) monthly concentrations by decade."}
knitr::include_graphics('figs/obsdat.png')
```

@@

p2a <- ggplot(toplo2, aes(x = SY, y = SYD)) + 
  geom_point(alpha = alph) + 
  geom_abline(slope = 1, intercept = 0, color = lncol) + 
  scale_y_log10(limits = ylm) + 
  scale_x_log10(limits = ylm) + 
  thm + 
  labs(
    title = '(b) Estimated chlorophyll between models'
  )
p2b <- ggplot(toplo2, aes(x = SYD, y = SYDI)) + 
  geom_point(alpha = alph) + 
  geom_abline(slope = 1, intercept = 0, color = lncol) + 
  scale_y_log10(limits = ylm) + 
  scale_x_log10(limits = ylm) + 
  thm
p2c <- ggplot(toplo2, aes(x = SY, y = SYDI)) + 
  geom_point(alpha = alph) + 
  geom_abline(slope = 1, intercept = 0, color = lncol) + 
  scale_y_log10(limits = ylm) + 
  scale_x_log10(limits = ylm) + 
  thm

@@

out <- p1 / p2 / p3
png('figs/modsumfig.png', height = 7, width = 8, units = 'in', family = 'serif', res = 400)
print(out)
dev.off()
```
```{r modsumfig, fig.cap = "GAM output of estimated chlorophyll for models SY, SYD, and SYDI.  Model S is identical to SY and is not shown.  Plots in (a) show model predictions by day of year with separate lines for each year.  Plots in (b) show pairwise comparisons of predicted chlorophyll between the models and plots in (c) show the same comparisons as in (b) but only for results from the estimated smoother for the `cont_year` variable.  The plots demonstrate that results between the models are comparable except for a few observations at extreme values(a), but they vary in the penalties applied to the basis functions for any particular smoother depending on which additive components are included in each model (b).  The 1:1 lines are in red to facilitate comparisons."}
knitr::include_graphics('figs/modsumfig.png')
```

@@

modobs <- rawdat %>% 
  filter(param == 'chl')
ylab <- 'Chlorophyll-a (ug/L)'

@@

ylim <- c(0, 21)
ylab <- expression(paste("Chlorophyll-a (", mu, "g ", L^-1, ")"))

@@

png('figs/trnddat.png', height = 6, width = 11, family = 'serif', units = 'in', res = 400)
print(p)
dev.off()
```
```{r trnddat, fig.cap = "Examples of seasonal averages and trend estimates using results of GAM predictions for station 34.  Plots (a), (b), and (c) show trend estimates for January through June and (d), (e), and (f) show trend estimates for July through December.  The trend lines estimate the rate of change of chlorophyll per year, reported as the log$_{10}$-slope (+/- 95 % confidence interval) in the plot title. ns: not significant at $\\alpha$ = 0.05, * p < 0.05, ** p < 0.005"}
knitr::include_graphics('figs/trnddat.png')
```

>>>>

Accurate and flexible trend assessment methods are valuable tools for describing historical changes in environmental monitoring datasets.  A key requirement is complete propagation of uncertainty through the analysis.  However, this is difficult when there are mismatches between time scales of monitoring data and analysis methods. Here, we propose a novel application of Generalized Additive Models (GAMs) to model seasonal and multi-decadal changes in a long-term monitoring dataset of chlorophyll-a concentrations in the San Francisco Estuary.  GAMs are a general method for modelling a response variable as the additive sum of one or more smoother splines fit to predictor variables.  These models have shown promise in other applications to water quality monitoring data to separate long-term (i.e., annual or decadal) trends from seasonal variation.  Our proposed methods estimate seasonal averages in a response variable with GAMs and then use the uncertainty measures with mixed-meta regression analyses to quantify inter-annual trends that account for full propagation of error across methods.  We first demonstrate that nearly identical descriptions of temporal changes can be obtained using different smoothing splines for annual or seasonal components of the time series.  We then extract seasonal averages for an a priori period of time within each year from the GAM results, including an accurate assessment of standard error within each year where the seasonal average is estimated.  Finally, we demonstrate how trends can be quantified with mixed-meta regression analyses that evaluate changes in the seasonal averages between years with full propagation of uncertainty in the estimates.  Overall, this approach leverages the ability of GAMs to identify an optimal amount of smoothing from the raw data and uses those results to accurately assess trends that minimize the likelihood of false positive or negative results, which is particularly important for time series with missing observations or varying sample effort over time.

@@

We evaluated near-surface chlorophyll (chl-a) data measured biweekly to monthly from 1990 to 2019 along the longitudinal axis of the SFE extending from Central Bay (stations 18-23), South Bay (stations 24-32), and Lower South Bay (stations 34-36) (Table \@ref(tab:sumtab), Figure \@ref(fig:sitemap)). Monitoring data were obtained from the SFE Research Program of the US Geological Survey [@Cloern16;@Schraga20].  Chl-a on GFF filters was analyzed fluorometrically after extraction in 90% acetone. Sampling frequency varied somewhat over time and by station.  Every observation was included directly in the statistical models without spatial or temporal binning or averaging.

@@

where `y` is the time-series of interest, such as chl-a, `cont_year` is "continuous year", a continuous numerical date (e.g., July 1st 2019 would be 2019.5), `y ~ s(...)` indicates that `y` will be explained by a smoothing spline (in this case of `cont_year`), and `num_knots_Y` is the number of knot or "connections" along the spline that influence curvature.  Log$_{10}$-transformed chl-a was used for all analyses to meet assumptions of normally-distributed residuals.

@@

Model SYD has the appealing feature that, if some parts of some years have limited data, model SYD will consider data from the same period in other years by imputing an average seasonal pattern with the `doy` spline. However, an interpretation of these imputations is challenging.  For example, the spring chl-a peak is a notable feature every year in SFE.  If the peak occurs at the same time every year but the magnitude varies, then the average within-year pattern can be interpreted as the average magnitude. However, if the magnitude is the same but the timing varies across years, then the magnitude of the average peak cannot be similarly interpreted and instead underestimates the magnitude that usually occurs.  Moreover, the width or duration of the peak will be longer than typically occurs in a given year.

@@

The distinct aspect of model SYDI is the anticipation that within-year fluctuations will vary smoothly from year to year.  For example, if one year has an especially early and large magnitude spring bloom, the years before and after would be expected also to have earlier and larger than average spring blooms. For the raw SFE data, and chl-a dynamics in many estuaries, this is not expected to be the case. An early and large spring bloom could easily be followed the next year by a late and relatively small spring bloom.  If the interaction spline has many knots, then the degree of smoothness across years can be estimated to be very low, allowing fluctuations within years to vary strongly across years.  However, if that is the case, the conceptual motivation for model SYDI may not be supported by the data. Given that observed bloom dynamics for most systems do not vary smoothly across years, the potential benefits of using model SYDI could be outweighed by using simpler models.  Moreover, the spline and penalty structure differ from the other models, making it difficult to evaluate their relative performance (Table \@ref(tab:modstrtab)).

@@

We visually compare chl-a estimates from models SY, SYD, and SYDI to emphasize that similar fits can be achieved by all of the presented models (Figure \@ref(fig:modsumfig), S is identical to SY and is not shown).  Models SY, SYD, and SYDI were fit to chl-a data from station 34 using large k values for the arguments `num_knots_y`, `num_knots_D`, `num_knots_Y_ti`, and `num_knots_D_ti` for each model.  Predictions by day of year from each model are visually similar (Figure \@ref(fig:modsumfig)a) and closely follow the 1:1 line (Figure \@ref(fig:modsumfig)b).  However, when contrasting the models using only the continuous year smoother (`s(cont_year)`, as an example), the model fits differ substantially because of structural differences and differences in the penalties applied to the basis functions.  These results are also reflected in differences in the effective degrees of freedom among the additive components of each model (Table \@ref(tab:modstrtab)). Accordingly, even though the models differ by which structural component describes variation in the chl-a time series, they provide similar predictions.

For all results, model S was used with enough knots in `num_knots_y` to evaluate chl-a trends across the monitoring stations in SFE.  This model was chosen because of the relatively quicker processing time to fit the model, while providing nearly identical explanatory power as compared to the other models (Table \@ref(tab:modstrtab)).

@@

In the second-stage analysis, we estimated a seasonal average, such as the mean spring chl-a concentrations and the uncertainty in each year.  We define $\mu_t$ as the average for the period of interest (e.g., seasonal average) in year $t$, $\hmt$ as an estimate of $\mu_t$, and $\hshmt$ as the estimated standard error of $\hmt$. The season includes $n$ days.  For simplicity, the following text omits subscript $t$.

Point estimates of response values for the fitted GAM take the form $\hby = \bX \hbb$, where $\hbb$ is the vector of parameter estimates and $\mathbf{X}$ is a model matrix of explanatory variables, including spline basis function values.  Vector $\hbb$ includes both fixed effect parameters and spline parameters, and $\bX$ contains columns corresponding to each.  For example, using model SY, if a point estimate for chl-a is needed for a single day, given as `dec_year` = $r$, then $\bX$ would have a row with $1$ in the first column (for the intercept parameter), $r$ (for the linear time trend) in the second column, and an evaluation of each spline basis function at $r$ in the remaining columns. The number of spline basis functions is related to the number of knots. Note that $r$ can be any time, not necessarily the time of an observation.

@@

In stage three of the analysis, we used a meta-analysis method to evaluate linear trends across years of within-year features such as seasonal water quality, characterized by the within-year means ($\hmt$) and their standard errors ($\hshmt$) that we estimated in stage two of the analysis.  In the examples, we considered two periods of interest for chl-a, January-July and August-December, which are relevant to phytoplankton bloom phenology in SFE.

This analysis provided a direct answer to the question: Is there a significant linear trend across a group of years in a seasonal average, where the time-scale of the trend is chosen by the investigator?  For example, is there a trend in the spring chl-a average from 1990 to 2000? This question can also be posed in a moving-window manner across a time-series (e.g., spring average trend from 1990-2000, 1991-2001, etc.). For all analyses, the response data of interest are $\hmt$, $t = 1, \ldots, N$, with their associated standard errors, $\hshmt$.  $N$ is the number of years of the study.

@@

This method was applied to a chosen sequence or "window" of years for estimating the linear trend.  This approach evaluated whether there was a significant linear trend and provided an estimated rate of change (slope) in chl-a over a chosen series of years. Using the meta-analysis considered statistical properties derived from all the data in the years of interest (i.e., the seasonal averages and their uncertainty).  Years with more data have smaller standard errors for $\hmt$ compared to years with less data. The meta-analysis explicitly includes these differences in uncertainty magnitudes.

@@

The above methods were applied to each station by evaluating changes in seasonal averages from January to June and July to December for approximate ten year blocks from 1991 to 2000, 2000 to 2010, and 2010 to 2019. As an extension of these methods, a chosen sequence of years was applied across the time series to assess periods of time within which seasonal trends could be significant or not (e.g., spring average trend from 1990-2000, 1991-2001, etc.).  Although the initial year sequence and seasonal period to evaluate is chosen by the analyst, applying the method in a moving window approach reduces some of the ambiguity around points in the time series when trends may be changing with reasonable certainty.  The moving window approach applies the meta-analysis model from left to right in the time series across the seasonal averages to obtain the slope and significance of the estimated trend.  We chose a centered ten-year window where the model estimates are based on results at equal half-window widths to the left and right of a given year. Although a left- or right-centered window could also be applied, we limit the analysis to a centered window to demonstrate the concept.

@@

where the back-transformed, expected value of the response variable $y$ (chl-a) is a function of the mean value $\mu$ in log$_{10}$-space and a dispersion estimate $\sigma^2$ from the model.  The dispersion is obtained directly from the fitted GAM object as an estimate of the variance of the residuals divided by the model degrees of freedom.

Slope estimates that were used to determine significance of a seasonal trend across years from the mixed-meta analysis were not back-transformed.  While it is possible to back-transform model predictions, estimates of the mean values, and endpoints of confidence intervals, it is not possible for slope estimates in log$_{10}$-space that can change in arithmetic space across the values of the independent variable.  As such, all results reporting slope estimates as rates of change in chl-a per year were reported at the scale of the model in log$_{10}$-space.

@@

Model predictions for chl-a trends across all stations had an average R-squared value equal to `r aver2`%  (Table \@ref(tab:modprftab)) and ranging from `r minr2`% (station `r minst`) to `r maxr2`% (station `r maxst`). GAM predictions from north to south on the longitudinal axis showed more pronounced annual and seasonal changes in chl-a towards the more southern stations (Figure \@ref(fig:prddat)). All the models suggested 1) increasing chl-a from 1990 until 2005 to 2010, followed by decreasing chl-a until the end of the record in 2019, 2) a spring chl-a peak, particularly at South Bay stations, and 3) a fall chl-a peak that was smaller than the spring peak.  The magnitude of the fall peak did not vary noticeably by location (Figure \@ref(fig:prddat)).

@@

To demonstrate the differences in results that can be obtained with the trend tests, estimates of linear trends across years for different seasons are shown for station 34 (Figure \@ref(fig:trnddat)). All plots show trends in seasonal averages within ten-year windows, whereas rows a-c show estimates from January to June and rows d-f show estimates from July to December.  The seasonal trend analyses showed that January to June chl-a increased (log$_{10}$ chl-a slope 0.03 $\mu$g L$^{-1}$ yr$^{-1}$, 0.01-0.06 95\% confidence interval) form 1991 to 2000, whereas a trend for the same period in July to December was not observed.  Chl-a also increased from 2000 to 2010, but only for July to December (log$_{10}$ slope 0.03, 0.01-0.05 95\% confidence interval).  Finally, chl-a decreased from 2010 to 2019 but only for July to December (log$_{10} chl-a slope -0.02, -0.04-0 95\% confidence interval).  Because the trends were confined to certain times of the year, the seasonal estimates provide additional information beyond coarser estimates that cover the entire year.

Temporal changes varied among regions of the Bay and were in fact estimated to be moving in opposite directions.  Figure \@ref(fig:trndmap) shows results from similar analyses as those in figure \@ref(fig:trnddat), but applied to all stations.  
Mixed-meta regression analyses applied to seasonal averages showed that increases (based on $p < 0.05$) for the January to June period were observed at stations 32, 34, and 36 from 1991 to 2000 and station 18 from 2000 to 2010, whereas chl-a decreased for at stations 30 and 32 from 2010 to 2019. For the July to December period, increases were observed at stations 24, 27, 30, and 32 from 1991 to 2000 and stations 18, 21, 22, and 34 from 2000 to 2010, whereas decreases were observed at stations 30, 32, and 34 from 2010 to 2019.

@@

Results from a ten-year moving window comparison of seasonal trends provided additional context on when significant changes were occurring at each station (Figure \@ref(fig:winchg)).  Trends were observed at all stations that followed a general pattern of increases early in the record followed by decreases later in the record.  Increases and decreases were observed in both the January to June and July to December seasonal periods, with some notable exceptions.  In particular, the most southern stations (32, 34, 36) had more trends in the July to December period.  Additionally, trends in recent years in both seasonal periods were not observed for the more northern stations. For both seasonal periods, a change from increasing to decreasing chl-a occurred at most stations around 2007.

@@

Applying the same comparison to all stations showed that different trend analysis methods provided conflicting information on the magnitude and significance of the seasonal chl-a changes in each decade (Figure \@ref(fig:trndcmp)).  The slope estimates from the OLS models applied to the observed data were understandably more variable than the slope estimates from the GAM averages and mixed-meta methods, with much larger slopes observed especially at the more southern stations in the January to March period.  Slope estimates from the OLS models applied to the averages from the GAMs as compared to the mixed-meta results were more similar, excluding some of the slope estimates for the southern stations.  Differences in significance of trends between the OLS models applied to the GAM averages and the mixed-meta analyses were also observed, reflecting the ability of the former to more accurately assess significance of trends by accounting for uncertainty in the average estimates.

@@

The trend assessment approach does not address causality in the observed trends.  In this regard, the approach is like other trend assessment methods where the focus is on understanding direction, magnitude, and confidence in changes in water quality variables over time. However, a logical follow-up question is "what" is driving the trend after the trend has been adequately described.  This information has obvious implications for management decisions on factors that influence water quality changes, e.g., wastewater treatment upgrades, large-scale climatic factors, or flow regulation practices.  An advantage of GAMs is their flexibility in including alternative predictors, such that the significance of a predictor or comparison of nested models with and without different predictors can provide evidence of which predictors are driving the observed trends [@Wood02;@Zuur09]. In such cases, considerations of model structure can have direct implications on conclusions regarding potential causality.  As noted above, model structure in describing the long-term trend component was irrelevant for describing long-term trends, although a distinction here is made in the objective of the analysis.  Our goal herein was to describe chl-a changes relative to time where the predictors were variations on a general theme (e.g., season vs. year), whereas using GAMs with different predictors assess potential causality is a different application with alternative goals.  Therefore, using our approach to assess causality will require considerations of model structure given how GAMs could be used to assess different questions.

Finally, the evaluation of trends for alternative water quality variables in addition to chl-a is a simple and logical extension of the methods proposed in this study.  The long-term monitoring program maintained by USGS includes multiple parameters in addition to chl-a that can provide additional context into broader water quality trends in South Bay [@Cloern16;@Schraga20]. These parameters include salinity, temperature, light attenuation, dissolved oxygen, suspended particulate matter, and dissolved inorganic nutrients, which collectively can be used to provide a broader understanding of potential eutrophication patterns or ecosystem shifts at seasonal and multi-decadal scales.  Chl-a measurements can also be used to estimate gross primary production to assess process rates that may be more indicative of system function [@Jassby02;@Cloern07].  The open-source R package developed for this manuscript can be used for these analyses to provide additional insight into potential drivers of water quality change in SFE and other estuarine systems.

@@

ylab <- expression(paste("Chl-a (", italic(mu), "g ", L^-1, ")"))

@@

p <- p1 + p2 + plot_layout(ncol = 1)
png('figs/obsdat.png', height = 7, width = 8, family = 'serif', units = 'in', res = 400)
print(p)
dev.off()
```
```{r obsdat, fig.cap = "Observed chl-a concentrations for all stations in central and south San Francisco Estuary (18-36, Figure \\@ref(fig:sitemap)), with (a) annual summer/fall concentrations (Aug - Dec) and (b) monthly concentrations by decade."}
knitr::include_graphics('figs/obsdat.png')
```

@@

p2a <- ggplot(toplo2, aes(x = SY, y = SYD)) + 
  geom_point(alpha = alph) + 
  geom_abline(slope = 1, intercept = 0, color = lncol) + 
  scale_y_log10(limits = ylm) + 
  scale_x_log10(limits = ylm) + 
  thm + 
  labs(
    title = '(b) Estimated chl-a between models'
  )
p2b <- ggplot(toplo2, aes(x = SYD, y = SYDI)) + 
  geom_point(alpha = alph) + 
  geom_abline(slope = 1, intercept = 0, color = lncol) + 
  scale_y_log10(limits = ylm) + 
  scale_x_log10(limits = ylm) + 
  thm
p2c <- ggplot(toplo2, aes(x = SY, y = SYDI)) + 
  geom_point(alpha = alph) + 
  geom_abline(slope = 1, intercept = 0, color = lncol) + 
  scale_y_log10(limits = ylm) + 
  scale_x_log10(limits = ylm) + 
  thm

@@

out <- p1 / p2 / p3
png('figs/modsumfig.png', height = 7, width = 8, units = 'in', family = 'serif', res = 400)
print(out)
dev.off()
```
```{r modsumfig, fig.cap = "GAM output of estimated chl-a for models SY, SYD, and SYDI.  Model S is identical to SY and is not shown.  Plots in (a) show model predictions by day of year with separate lines for each year.  Plots in (b) show pairwise comparisons of predicted chl-a between the models and plots in (c) show the same comparisons as in (b) but only for results from the estimated smoother for the `cont_year` variable.  The plots demonstrate that results between the models are comparable except for a few observations at extreme values(a), but they vary in the penalties applied to the basis functions for any particular smoother depending on which additive components are included in each model (b).  The 1:1 lines are in red to facilitate comparisons."}
knitr::include_graphics('figs/modsumfig.png')
```

@@

modobs <- rawdat %>% 
  filter(param == 'chl')
ylab <- expression(paste("Chl-a (", mu, "g ", L^-1, ")"))

@@

ylim <- c(0, 21)
ylab <- expression(paste("Chl-a (", mu, "g ", L^-1, ")"))

@@

png('figs/trnddat.png', height = 6, width = 11, family = 'serif', units = 'in', res = 400)
print(p)
dev.off()
```
```{r trnddat, fig.cap = "Examples of seasonal averages and trend estimates using results of GAM predictions for station 34.  Plots (a), (b), and (c) show trend estimates for January through June and (d), (e), and (f) show trend estimates for July through December.  The trend lines estimate the rate of change of chl-a per year, reported as the log$_{10}$-slope (+/- 95 % confidence interval) in the plot title. ns: not significant at $\\alpha$ = 0.05, * p < 0.05, ** p < 0.005"}
knitr::include_graphics('figs/trnddat.png')
```
